NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2019-11-18 Manuscript/manuscriptPGPE.tex
Update calcium binding section

\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}

@@

\clearpage
\subsection{Cation binding to PE and PG lipid bilayers}
The headgroup order parameters of PC lipids can used to measure ion binding
affinity to lipid bilayers, because their magnitude is linearly proportional
to the amount of bound charge in bilayer according to the molecular
electrometer concept \cite{seelig87,catte16}. The molecular electrometer concept
can be used also for bilayers containing PC lipids mixed with charged
lipids \cite{borle85,macdonald87,roux90,antila19}.
%This is demonstrated in Figs \ref{OrderParametersWithCaCl},
%\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
%showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
%as a function of CaCl$_2$ concentration in the presence of different amounts of
%negatively charged PS or PG lipids.
%
%PC headgroup order parameters increase when negatively charged
%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
%as expected based on electrometer concept \cite{seelig87}
%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%Further, the order parameters decrease with the addition
%of CaCl$_2$ and the decrease is more pronounced for systems with more
%negatively charged lipids (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
%At CaCl$_2$ concentrations ($\sim$ 50-300mM) where order parameters reach the values for pure PC,
%the Ca2+ binding presumably fully cancels the charge from negative lipids and
%overcharging occurs above these concenterations.
The electrometer concept has been very useful in evaluating ion binding affinity in simulations
against experiments, because the headgroup order parameter changes as a function of ion concentration
can be directly compared with experiments \cite{catte16,melcr18,antila19}.

Calcium binding affinity to PC and PS lipid bilayers was not correctly described by
any of the standard MD simulation forced fields \cite{catte16,antila19}, while
recently introduced force field with electronic continuum correction (ECC) performed better \cite{melcr18}.
The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
but the decrease of $\beta$-carbon order parameter seems to be overestimated.
However, the $\beta$-carbon order parameter was not actually measured from these samples,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
However, when simulation was continued for $\mu$s, the binding affinity substantially increased
and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
conclusion seems to be similar here, although the new NBfix parameters may complicate
the situation \todo{The status of NBfix parameters in these simulations should be checked.}.
%Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
%suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
%POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
%parameter of bound charge
%\todo{The response of CHARMM36 to cationic surfactant against experiments \cite{scherer89} to be checked.
%  I have already ran the simulations, analysis to be done.}.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG.eps}
  \caption{\label{changesWITHCaClPG}
    (left) The headgroup order parameters of PC from PC:PG mixtures as a function CaCl$_2$
    concentration from experiments \cite{borle85,macdonald87} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured.
    (right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}

The $\beta$-carbon order parameter of PG exhibits a rapid decrease with small CaCl$_2$ concentrations
and a more modest decrease with larger concentrations in experiments \cite{borle85} (Fig. \ref{PSPGchangesWITHCaCl}).
The rapid decrease with CaCl$_2$ is observed but overestimated in CHARMM36 simulation
with POPC:POPG 1:1 mixture, but not in 4:1 mixture \todo{This is little bit weird, should be checked.}.

@@

\todo{We need PC:PG simulations with CaCl$_2$ from different force fields to finish the discussion.}

>>>>

\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
%\affiliation{Institute of Organic Chemistry and Biochemistry,
%Academy of Sciences of the Czech Republic, 
%Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}

@@

\clearpage
\subsection{Calcium binding to PE and PG lipid bilayers}
The headgroup order parameters of PC lipids can used to measure ion binding
affinity to lipid bilayers, because their magnitude is linearly proportional
to the amount of bound charge in bilayer according to the molecular
electrometer concept \cite{seelig87,catte16}. The molecular electrometer concept
can be used also for bilayers containing PC lipids mixed with charged
lipids \cite{borle85,macdonald87,roux90,antila19}.
The electrometer concept has been very useful in evaluating ion binding affinity in simulations
against experiments, because the headgroup order parameter changes as a function of ion concentration
can be directly compared with experiments \cite{catte16,melcr18,antila19}.

Calcium binding affinity to PC and PS lipid bilayers was not correctly described by
any of the standard MD simulation forced fields \cite{catte16,antila19}, while
recently introduced force field with electronic continuum correction (ECC) performed better \cite{melcr18}.
The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
but the decrease of $\beta$-carbon order parameter seems to be overestimated.
However, the $\beta$-carbon order parameter was not actually measured from these samples,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
On the other hand, the simulations are not converged.
\todo{We still need more data to finish the discussion. More detailed discussion is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/12}
% The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
%However, when simulation was continued for $\mu$s, the binding affinity substantially increased
%and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
%conclusion seems to be similar here, although the new NBfix parameters may complicate
%the situation \todo{The status of NBfix parameters in these simulations should be checked.}.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG.eps}
  \caption{\label{changesWITHCaClPG}
    (left) The headgroup order parameters of PC from PC:PG mixtures as a function CaCl$_2$
    concentration from experiments \cite{borle85,macdonald87} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured.
    (right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}

%The $\beta$-carbon order parameter of PG exhibits a rapid decrease with small CaCl$_2$ concentrations
%and a more modest decrease with larger concentrations in experiments \cite{borle85} (Fig. \ref{PSPGchangesWITHCaCl}).
%The rapid decrease with CaCl$_2$ is observed but overestimated in CHARMM36 simulation
%with POPC:POPG 1:1 mixture, but not in 4:1 mixture \todo{This is little bit weird, should be checked.}.

@@

%\todo{We need PC:PG simulations with CaCl$_2$ from different force fields to finish the discussion.}
